---
title: "Cell chat interaction analysis, case `r params$PROJECT`"
author:
  - "trnguyen@ukaachen.de"
date: "Last update on `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    keep_md: no
    df_print: paged
    toc: true
    toc_float:
      toc_collapsed: false
    toc_depth: 3
    number_sections: true
    theme: lumen
params:
  PROJECT: NA
---
  
  
```{css zoom-lib-src, echo = FALSE}
script src = "https://ajax.googleapis.com/ajax/libs/jquery/3.4.1/jquery.min.js"
```

```{js zoom-jquery, echo = FALSE}
$(document).ready(function() {
  $('body').prepend('<div class=\"zoomDiv\"><img src=\"\" class=\"zoomImg\"></div>');
  // onClick function for all plots (img's)
    $('img:not(.zoomImg)').click(function() {
      $('.zoomImg').attr('src', $(this).attr('src')).css({width: '100%'});
      $('.zoomDiv').css({opacity: '1', width: 'auto', border: '1px solid white', borderRadius: '5px', position: 'fixed', top: '50%', left: '50%', marginRight: '-50%', transform: 'translate(-50%, -50%)', boxShadow: '0px 0px 50px #888888', zIndex: '50', overflow: 'auto', maxHeight: '100%'});
    });
    // onClick function for zoomImg
    $('img.zoomImg').click(function() {
      $('.zoomDiv').css({opacity: '0', width: '0%'}); 
    });
  });
```

<style type="text/css">
    div.datatables { height: auto !important;}
</style>

```{r echo=FALSE, warning=FALSE, message=FALSE, results='hide', fig.height=10, fig.width=18}
##### clean up #####
# gc()
# rm(list = ls())

path.to.project.src <- "/home/hieunguyen/CRC1382/src_2023/SBharadwaj/deep_seq_batch"

source(file.path(path.to.project.src, "00_import_libraries.R"))
source(file.path(path.to.project.src, "00_helper_functions.R"))

#####----------------------------------------------------------------------#####
# CONFIGURATIONS AND PREPRATIONS
#####----------------------------------------------------------------------#####
outdir <- "/home/hieunguyen/CRC1382/outdir"
# PROJECT <- params$PROJECT
PROJECT <- "SBharadwaj_20240318_Sample_1_4"

path.to.main.input <- file.path(outdir, "SeuratV5", PROJECT)
path.to.main.output <- file.path(path.to.main.input, "data_analysis")
dir.create(path.to.main.output, showWarnings = FALSE, recursive = TRUE)

path.to.01.output <- file.path(path.to.main.output, "01_output")
path.to.02.output <- file.path(path.to.main.output, "02_output")
path.to.04.output <- file.path(path.to.main.output, "04_output")
dir.create(path.to.04.output, showWarnings = FALSE, recursive = TRUE)

s.obj <- readRDS(file.path(path.to.01.output,sprintf("Project_%s.rds", PROJECT)))

meta.data <- s.obj@meta.data %>% rownames_to_column("barcode") %>% rowwise() %>%
  mutate(condition = str_split(name, "_")[[1]][[1]]) %>%
  mutate(organ = str_split(name, "_")[[1]][[2]]) %>%
  mutate(celltype = str_split(name, "_")[[1]][[3]])  %>% 
  column_to_rownames("barcode")

meta.data <- meta.data[row.names(s.obj@meta.data),]

s.obj <- AddMetaData(object = s.obj, metadata = meta.data$condition, col.name = "condition")
s.obj <- AddMetaData(object = s.obj, metadata = meta.data$organ, col.name = "organ")
s.obj <- AddMetaData(object = s.obj, metadata = meta.data$celltype, col.name = "celltype")

```

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.height=10, fig.width=14}
DimPlot(objec = s.obj, reduction = "cca_UMAP", label = TRUE, label.box = TRUE, repel = TRUE)
```

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.height=10, fig.width=14}
DimPlot(objec = s.obj, reduction = "cca_UMAP", label = TRUE, label.box = TRUE, repel = TRUE, group.by = "name")
```

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.height=10, fig.width=14}
DimPlot(objec = s.obj, reduction = "cca_UMAP", label = TRUE, label.box = TRUE, repel = TRUE, group.by = "prediction2")
```

## All samples in integrated UMAP {.tabset}
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
Idents(s.obj) <- "cca.cluster.0.5"
FeaturePlot(s.obj, features = c("Ciita", "Foxp3", "Klrg1", "Gata3"), label = TRUE)
```

## Each sample in integrated UMAP {.tabset}
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
Idents(s.obj) <- "cca.cluster.0.5"
for (sample.id in unique(s.obj$name)){
  cat(sprintf("### Sample %s \n", sample.id))
  p <- FeaturePlot(subset(s.obj, name == sample.id), features = c("Ciita", "Foxp3", "Klrg1", "Gata3"), label = TRUE)  
  print(p)
  cat("\n \n")
}
```

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
##### Assign cells to groups by the expression of Ciita, Foxp3, Klrg1 and Gata3 genes
cells <- hash()
for (input.gene in c("Ciita", "Foxp3", "Klrg1", "Gata3")){
  tmp <- GetAssayData(s.obj, assay = "SCT", slot = "data")[input.gene, ]  
  cells[[input.gene]] <- tmp[tmp > 0] %>% names()
}

group1.1 <- intersect(cells$Foxp3, cells$Klrg1)
group1.2 <- intersect(cells$Foxp3, cells$Gata3)

group1.cells <- unique(c(group1.1, group1.2))
group2.cells <- cells$Ciita

meta.data <- s.obj@meta.data %>% rownames_to_column("barcode") %>%
  rowwise() %>%
  mutate(group = ifelse(barcode %in% group1.cells, "Group1", ifelse(barcode %in% group2.cells, "Group2", "others"))) %>%
  column_to_rownames("barcode")

meta.data <- meta.data[row.names(s.obj@meta.data), ]

s.obj <- AddMetaData(object = s.obj, metadata = meta.data$group, col.name = "group")

```

## UMAP: groups
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
FeaturePlot(
  object = s.obj, reduction = "cca_UMAP", features = c("Ciita", "Foxp3", "Klrg1", "Gata3"), ncol = 2
)
```


```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
DimPlot(object = s.obj, reduction = "cca_UMAP", label = TRUE, label.box = TRUE, repel = TRUE, group.by = "group", pt.size = 1)
```

## Violin plot: Ciita, Foxp3, Klrg1, Gata3 genes, all samples

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
VlnPlot(object = s.obj, features = c("Ciita", "Foxp3", "Klrg1", "Gata3"), group.by = "cca.cluster.0.5", ncol = 2, assay = "SCT", slot = "data")
```

## Violin plot: Ciita, Foxp3, Klrg1, Gata3 genes, each sample {.tabset}
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
for (sample.id in unique(s.obj$name)){
  cat(sprintf("### Sample %s \n", sample.id))
  p <- VlnPlot(object = subset(s.obj, name == sample.id), features = c("Ciita", "Foxp3", "Klrg1", "Gata3"), group.by = "cca.cluster.0.5", ncol = 2, assay = "SCT", slot = "data")  
  print(p)
  cat("\n \n")
}
```


```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
group1.clusters <- unique(intersect(predefined.clusters[[PROJECT]][["Foxp3"]], predefined.clusters[[PROJECT]][["Klrg1"]]),
                          intersect(predefined.clusters[[PROJECT]][["Foxp3"]], predefined.clusters[[PROJECT]][["Gata3"]]))
group2.clusters <- predefined.clusters[[PROJECT]][["Ciita"]]

group1.clusters <- to_vec( for (item in group1.clusters) sprintf("cluster%s", item))
group2.clusters <- to_vec( for (item in group2.clusters) sprintf("cluster%s", item))

meta.data <- s.obj@meta.data %>% rownames_to_column("barcode") %>%
  mutate(clusterid = sprintf("cluster%s", cca.cluster.0.5)) %>%
  column_to_rownames("barcode")

meta.data$clusterid <- factor(meta.data$clusterid, levels = to_vec( for(item in seq(0, length(unique(s.obj$cca.cluster.0.5)) - 1))sprintf("cluster%s", item))) 
  
meta.data <- meta.data[row.names(s.obj@meta.data),]
s.obj <- AddMetaData(object = s.obj, metadata = meta.data$clusterid, col.name = "clusterid")
```


```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
print("List of clusters that we focus on in CellChat results: \n")
print(sprintf("Foxp3 and Klrg1 OR Foxp3 and Gata3: %s \n", paste(group1.clusters, collapse = ", ")))

print(sprintf("Ciita: %s", paste(group2.clusters, collapse = ", ")))
```


```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
```

## Number of cells in each assigned cluster ID
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
table(s.obj$cca.cluster.0.5) %>% as.data.frame() %>% arrange(desc(Freq))
```

# Cell-cell communication inference with CellChat

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
if ("CellChat" %in% installed.packages() == FALSE){
  devtools::install_github("jinworks/CellChat")
  remove.packages("Matrix")
  install.packages("/media/hieunguyen/HD0/storage/offline_pkgs/Matrix_1.5-4.1.tar.gz", type = "source", repos = NULL)
    install.packages("/media/hieunguyen/HD0/storage/offline_pkgs/ggplot2_3.4.4.tar.gz", type = "source", repos = NULL)
}
library(CellChat)

if (file.exists(file.path(path.to.04.output, sprintf("%s.cellchat.cca.cluster.0.5.rds", PROJECT))) == FALSE){
  data.input <- s.obj[["SCT"]]@data 
  Idents(s.obj) <- "clusterid"
  labels <- Idents(s.obj)
  meta <- data.frame(labels = labels, row.names = names(labels)) 
  cellchat <- createCellChat(object = s.obj, group.by = "ident", assay = "SCT")
  
  ##### Select the CellChat database: human or mouse
  CellChatDB <- CellChatDB.mouse 
  showDatabaseCategory(CellChatDB)
  
  CellChatDB.use <- subsetDB(CellChatDB) # use all, except non protein signaling
  cellchat@DB <- CellChatDB.use
  
  ##### data in cellchat object pre-processing
  cellchat <- subsetData(cellchat) # This step is necessary even if using the whole database
  future::plan("multisession", workers = 4) # do parallel
  if ("presto" %in% installed.packages() == FALSE){
    devtools::install_github('immunogenomics/presto')
  }
  
  cellchat <- identifyOverExpressedGenes(cellchat)
  cellchat <- identifyOverExpressedInteractions(cellchat)
  
  ##### Compute the communication prob
  cellchat <- computeCommunProb(cellchat, type = "triMean")
  cellchat <- filterCommunication(cellchat, min.cells = 10)
  
  ##### infer cell-cell communication at a signaling pathway
  cellchat <- computeCommunProbPathway(cellchat)
  
  ##### aggregate cell-cell communication network
  cellchat <- aggregateNet(cellchat)
  
  saveRDS(object = cellchat, file.path(path.to.04.output, sprintf("%s.cellchat.cca.cluster.0.5.rds", PROJECT)))  
} else {
  cellchat <- readRDS(file.path(path.to.04.output, sprintf("%s.cellchat.cca.cluster.0.5.rds", PROJECT)))
}
```

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
groupSize <- as.numeric(table(cellchat@idents))
names(groupSize) <- unlist(lapply(
  seq(1, length(unique(cellchat@idents))), function(x){
    return(sprintf("cluster%s", as.character(x)))
  })
)
```


```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
par(mfrow = c(1,2), xpd=TRUE)
netVisual_circle(cellchat@net$count[c(group1.clusters, group2.clusters), 
                            c(group1.clusters, group2.clusters)], 
                 vertex.weight = groupSize[c(group1.clusters, group2.clusters)], weight.scale = T, label.edge= F, title.name = "Number of interactions")
```

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
netVisual_circle(cellchat@net$weight[c(group1.clusters, group2.clusters), 
                            c(group1.clusters, group2.clusters)], 
                 vertex.weight = groupSize[c(group1.clusters, group2.clusters)], weight.scale = T, label.edge= F, title.name = "Interaction weights/strength")
```

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
mat <- cellchat@net$weight
par(mfrow = c(3,4), xpd=TRUE)
for (i in 1:nrow(mat)) {
  mat2 <- matrix(0, nrow = nrow(mat), ncol = ncol(mat), dimnames = dimnames(mat))
  mat2[i, ] <- mat[i, ]
  netVisual_circle(mat2, vertex.weight = groupSize, weight.scale = T, edge.weight.max = max(mat), title.name = rownames(mat)[i])
}
```

## Table: Signal between each pair of clusters
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
mat.weight <- cellchat@net$weight
mat.count <- cellchat@net$count

matP.weight <- cellchat@netP$prob
```

### Counts
The number of interactions between any two cell groups  (row --> column, row: source, column: target)
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
mat.count %>% create_dt()
```

### Weights
The total reaction strength - weights. Weights are proportional to the interaction strength.
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
mat.weight %>% create_dt()
```

# Visualization: interactions
## Aggregated interaction network between all clusters {.tabset}
### The number of interactions (counts)
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
par(mfrow = c(1,2), xpd=TRUE)
netVisual_circle(mat.count[c(group1.clusters, group2.clusters), 
                            c(group1.clusters, group2.clusters)], 
                 vertex.weight = groupSize[c(group1.clusters, group2.clusters)], weight.scale = T, label.edge= F, title.name = "Number of interactions")
```

### The strength of interactions (weights)
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
netVisual_circle(mat.weight[c(group1.clusters, group2.clusters), 
                            c(group1.clusters, group2.clusters)], 
                 vertex.weight = groupSize[c(group1.clusters, group2.clusters)], 
                 weight.scale = T, 
                 label.edge= F, 
                 title.name = "Interaction weights/strength")
```

## Sent signals from each cell group
Since it is difficult to interpret the interactions on the above aggregated plots, here we show the strength of signals sent from each cluster to the others. Thicker edge line indicates a stronger signal. In the Circle plot below, circle sizes are proportional to the number of cells in each cluster. 

The weights can be compared globally between clusters; for example, if the line connecting cluster 0 and cluster 10 in the plot `cluster 0` is **thicker** than the line connecting cluster 1 and cluster 10, we can say that the sending signal from 0 --> 10 is **stronger** than 1 --> 10.  

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
par(mfrow = c(2, 2), xpd=TRUE)
for (i in 1:nrow(mat.weight)) {
  mat2 <- matrix(0, nrow = nrow(mat.weight), ncol = ncol(mat.weight), dimnames = dimnames(mat.weight))
  mat2[i, ] <- mat.weight[i, ]
  mat2[i, i] <- 0
  strength <- mat2[i, ] %>% sort() %>% tail(5)
  strength <- strength[strength != 0]
  names(strength) <- to_vec(for (item in names(strength)) str_replace(item, "cluster_", ""))
  title <- sprintf("Top 5 highest signals from cluster %s: %s", rownames(mat.weight)[i], paste(names(strength), collapse = ", "))
  netVisual_circle(mat2, vertex.weight = groupSize, weight.scale = T, edge.weight.max = max(mat.weight), title.name = title, arrow.size = 1)
}
```

## Received signals from each cell group

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
par(mfrow = c(2, 2), xpd=TRUE)
for (i in 1:nrow(mat.weight)) {
  mat2 <- matrix(0, nrow = nrow(mat.weight), ncol = ncol(mat.weight), dimnames = dimnames(mat.weight))
  mat2[, i] <- mat.weight[, i]
  mat2[i, i] <- 0
  strength <- mat2[, i] %>% sort() %>% tail(5) 
  strength <- strength[strength != 0]
  names(strength) <- to_vec(for (item in names(strength)) str_replace(item, "cluster_", ""))
  title <- sprintf("Top 5 highest signals to cluster %s: %s", rownames(mat.weight)[i], paste(names(strength), collapse = ", "))
  netVisual_circle(mat2, vertex.weight = groupSize, weight.scale = T, edge.weight.max = max(mat.weight), title.name = title, arrow.size = 1)
}

```

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
all.pathways <- cellchat@netP$pathways
sprintf("Total number of pathways: %s", length(all.pathways))
print(all.pathways)
```


```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
groups <- split(all.pathways, seq(1, 7))
```


```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
interaction.on.pathways <- hash()
for (pathway in all.pathways){
  interaction.on.pathways[[pathway]] <- cellchat@netP$prob[, , pathway]
}
```

# Visualization on pathway levels, sent signals 

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
all.clusters <- row.names(mat.weight)
for (group in names(groups)){
  cat(sprintf("## Group %s  \n \n", group))
  cat("\n \n")
  for (pathway in groups[[group]]){
    cat(sprintf("### Pathway: %s {.tabset} \n", pathway))
    for (cluster.id in c(group1.clusters, group2.clusters)){
      par(mfrow = c(1, 1))
      cat(sprintf("#### Cluster %s \n", cluster.id))
      netVisual_aggregate(cellchat, signaling = c(pathway),
                        layout = "circle", 
                        sources.use = c(cluster.id), 
                        targets.use = setdiff(row.names(mat.weight), c(cluster.id)),
                        arrow.size = 1) 
      cat("\n \n")
    }
  }
}
```

# Visualization on pathway levels, received signals 

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
all.clusters <- row.names(mat.weight)
for (group in names(groups)){
  cat(sprintf("## Group %s  \n \n", group))
  cat("\n \n")
  for (pathway in groups[[group]]){
    cat(sprintf("### Pathway: %s {.tabset} \n", pathway))
    for (cluster.id in c(group1.clusters, group2.clusters)){
      par(mfrow = c(1, 1))
      cat(sprintf("#### Cluster %s \n", cluster.id))
      netVisual_aggregate(cellchat, signaling = c(pathway),
                        layout = "circle", 
                        targets.use = c(cluster.id), 
                        sources.use = setdiff(row.names(mat.weight), c(cluster.id)),
                        arrow.size = 1) 
      cat("\n \n")
    }
  }
}
```

# Heatmap
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
for (group in names(groups)){
  cat(sprintf("## Group %s {.tabset} \n \n", group))
  cat("\n \n")
  for (pathway in groups[[group]]){
    cat(sprintf("### Pathway: %s \n", pathway))
    tryCatch({
      p <- netVisual_heatmap(cellchat, 
                           signaling = c(pathway),
                           color.heatmap = "Reds",
                           slot.name = "netP")
  }, error=function(e){
    p <- ggplot() + ggtitle("No data to show")
  })
    print(p)
    cat("\n \n")
  }
}

```

