---
title: "Data analysis for sub-clustering data, Project `r params$PROJECT`, sub-clustering version `r params$sub.cluster.idx`"
author:
  - "trnguyen@ukaachen.de"
date: "Last update on `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    keep_md: no
    df_print: paged
    toc: true
    toc_float:
      toc_collapsed: false
    toc_depth: 3
    number_sections: true
    theme: lumen
params:
  PROJECT: NA
  sub.cluster.idx: NA
---
  
  
```{css zoom-lib-src, echo = FALSE}
script src = "https://ajax.googleapis.com/ajax/libs/jquery/3.4.1/jquery.min.js"
```

```{js zoom-jquery, echo = FALSE}
$(document).ready(function() {
  $('body').prepend('<div class=\"zoomDiv\"><img src=\"\" class=\"zoomImg\"></div>');
  // onClick function for all plots (img's)
    $('img:not(.zoomImg)').click(function() {
      $('.zoomImg').attr('src', $(this).attr('src')).css({width: '100%'});
      $('.zoomDiv').css({opacity: '1', width: 'auto', border: '1px solid white', borderRadius: '5px', position: 'fixed', top: '50%', left: '50%', marginRight: '-50%', transform: 'translate(-50%, -50%)', boxShadow: '0px 0px 50px #888888', zIndex: '50', overflow: 'auto', maxHeight: '100%'});
    });
    // onClick function for zoomImg
    $('img.zoomImg').click(function() {
      $('.zoomDiv').css({opacity: '0', width: '0%'}); 
    });
  });
```

<style type="text/css">
    div.datatables { height: auto !important;}
</style>

**Note on the .html report: Click on the image to zoom-in and one more click to go back to the original**

```{r echo=FALSE, warning=FALSE, message=FALSE, results='hide', fig.height=10, fig.width=18}
##### clean up #####
# gc()
# rm(list = ls())

path.to.project.src <- "/home/hieunguyen/CRC1382/src_2023/SBharadwaj/deep_seq_batch"

source(file.path(path.to.project.src, "00_import_libraries.R"))
source(file.path(path.to.project.src, "00_helper_functions.R"))

#####----------------------------------------------------------------------#####
# CONFIGURATIONS AND PREPRATIONS
#####----------------------------------------------------------------------#####

path.to.project.src <- "/home/hieunguyen/CRC1382/src_2023/SBharadwaj/deep_seq_batch"
scrna_pipeline_src <- "/home/hieunguyen/CRC1382/src_2023/src_pipeline/scRNA_GEX_pipeline_SeuratV5/processes_src"

source(file.path(path.to.project.src, "00_import_libraries.R"))
source(file.path(path.to.project.src, "00_helper_functions.R"))
source(file.path(path.to.project.src, "sub_clustering_indices.R"))
source(file.path(scrna_pipeline_src, "s8_integration_and_clustering_SeuratV5.R"))

#####----------------------------------------------------------------------#####
# CONFIGURATIONS AND PREPRATIONS
#####----------------------------------------------------------------------#####
outdir <- "/media/hieunguyen/HD01/outdir/CRC1382/SBharadwaj_20240318"
PROJECT <- params$PROJECT
sub.cluster.idx <- params$sub.cluster.idx

path.to.main.input <- file.path(outdir, PROJECT)
path.to.main.output <- file.path(path.to.main.input, "data_analysis_with_cell_annotations")
path.to.06.output <- file.path(path.to.main.output, "06_output", sub.cluster.idx)
dir.create(path.to.06.output, showWarnings = FALSE, recursive = TRUE)

path.to.save.subclusters <- file.path(path.to.main.output, "sub_clusters", sub.cluster.idx)

s.obj <- readRDS(file.path(path.to.save.subclusters, "s8_output", sprintf("%s.output.s8.rds", PROJECT)))
```

# Dimension reduction with UMAP

## UMAP: all clusters 
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
DimPlot(object = s.obj, reduction = "cca_UMAP", label = TRUE, label.box = TRUE, label.size = 10, pt.size = 1, repel = TRUE, group.by = "cell.annotation") + 
  ggtitle(sprintf("UMAP: All clusters")) +
  theme(axis.text = element_text(size=20, face = "bold"),
        axis.title = element_text(size=20, face = "bold"), 
        title = element_text(size = 20, face = "bold"))
```

## UMAP: all samples
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
DimPlot(object = s.obj, reduction = "cca_UMAP", label = TRUE, label.box = TRUE, label.size = 10, pt.size = 1, repel = TRUE, group.by = "name") + 
  ggtitle(sprintf("UMAP Sample: All samples after integrating")) +
  theme(axis.text = element_text(size=20, face = "bold"),
        axis.title = element_text(size=20, face = "bold"), 
        title = element_text(size = 20, face = "bold"))
```



# Cluster marker genes

```{r echo=FALSE, warning=FALSE, message=FALSE, results='hide', fig.height=10, fig.width=18}
Idents(s.obj) <- "cell.annotation"
if (file.exists(file.path(path.to.06.output, "DE_cluster_marker_genes.rds")) == FALSE){
  DefaultAssay(s.obj) <- "SCT"
  s.obj <- PrepSCTFindMarkers(s.obj)
  cluster.markers <- FindAllMarkers(object = s.obj, assay = "SCT", test.use = "wilcox", slot = "data", min.pct = 0.5)
  cluster.markers <- subset(cluster.markers, cluster.markers$p_val_adj < 0.05 & cluster.markers$avg_log2FC > 0)
  saveRDS(cluster.markers, file.path(path.to.06.output, "DE_cluster_marker_genes.rds"))
} else {
  cluster.markers <- readRDS(file.path(path.to.06.output, "DE_cluster_marker_genes.rds"))
}
```

Identify differentially expressed genes in each cluster. 

## Feature plot {.tabset}
```{r echo=FALSE, fig.height=12, fig.width=20, message=FALSE, warning=FALSE, results='asis'}
for (cluster.id in sort(unique(cluster.markers$cluster))){
  DefaultAssay(s.obj) <- "SCT"
  cat(sprintf("### Cluster %s \n", cluster.id))
  tmp.cluster.markers <- subset(cluster.markers, cluster.markers$cluster == cluster.id) %>% arrange(desc(avg_log2FC))
  p <- FeaturePlot(object = s.obj, reduction = "cca_UMAP", features = head(tmp.cluster.markers, 9)$gene, ncol = 3, label = TRUE, pt.size = 0.5, label.size = 5, label.color = "red", slot = "data")  
  print(p)
  cat("\n \n")
}
```

## Dot plot{.tabset}
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
for (cluster.id in sort(unique(cluster.markers$cluster))){
  DefaultAssay(s.obj) <- "SCT"
  cat(sprintf("### Cluster %s \n", cluster.id))
  tmp.cluster.markers <- subset(cluster.markers, cluster.markers$cluster == cluster.id) %>% arrange(desc(avg_log2FC))
  p <- DotPlot(object = s.obj, features = head(tmp.cluster.markers, 9)$gene)  
  print(p)
  cat("\n \n")
}
```

## Violin plot{.tabset}
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
for (cluster.id in sort(unique(cluster.markers$cluster))){
  DefaultAssay(s.obj) <- "SCT"
  cat(sprintf("### Cluster %s \n", cluster.id))
  tmp.cluster.markers <- subset(cluster.markers, cluster.markers$cluster == cluster.id) %>% arrange(desc(avg_log2FC))
  p <- VlnPlot(object = s.obj, features = head(tmp.cluster.markers, 9)$gene, slot = "data")  
  print(p)
  cat("\n \n")
}
```

## Full tables of DE genes {.tabset}
```{r echo=FALSE, results='asis', include=FALSE}
cluster.markers %>% create_dt()
```

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
all.cluster.ids <- sort(unique(s.obj$cell.annotation))
for (cluster.id in all.cluster.ids){
  tmp.table <- subset(cluster.markers, cluster.markers$cluster == cluster.id) %>% arrange(desc(avg_log2FC))
  cat(paste("\n\n### Cluster: ", cluster.id, "##\n"))
  print( htmltools::tagList(datatable(tmp.table, extensions = 'Buttons',
                                      filter = "top",
                                      options = list(dom = 'Blfrtip',
                                                     buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
                                                     lengthMenu = list(c(10,25,50,-1),
                                                                       c(10,25,50,"All")),
                                                     columnDefs = list(list(
                                                       targets = "_all",
                                                       render = JS(
                                                         "function(data, type, row, meta) {",
                                                         "return type === 'display' && data != null && data.length > 100 ?",
                                                         "'<span title=\"' + data + '\">' + data.substr(0, 100) + '...</span>' : data;",
                                                         "}")
                                                     ))
                                      ))))
  cat("\n \n")  
}
```

# Number of cells in each cluster in each sample
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
cell.count.in.clusters <- table(s.obj$name, s.obj$cell.annotation) %>% as.data.frame() %>%
  pivot_wider(names_from = "Var2", values_from = "Freq") 
cell.count.in.clusters %>% create_dt()
```

# UMAP: expression of some genes

## All samples in integrated UMAP {.tabset}
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
Idents(s.obj) <- "cell.annotation"
FeaturePlot(s.obj, features = c("Ciita", "Foxp3", "Klrg1", "Gata3"), label = TRUE, reduction = "cca_UMAP")
```

## Each sample in integrated UMAP {,tabset}
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
Idents(s.obj) <- "cell.annotation"
for (sample.id in unique(s.obj$name)){
  cat(sprintf("### Sample %s \n", sample.id))
  p <- FeaturePlot(subset(s.obj, name == sample.id), features = c("Ciita", "Foxp3", "Klrg1", "Gata3"), label = TRUE, reduction = "cca_UMAP")  
  print(p)
  cat("\n \n")
}

```

## Violin plot: Ciita, Foxp3, Klrg1, Gata3 genes, all samples

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
VlnPlot(object = s.obj, features = c("Ciita", "Foxp3", "Klrg1", "Gata3"), group.by = "cell.annotation", ncol = 2, assay = "SCT", slot = "data")
```

## Violin plot: Ciita, Foxp3, Klrg1, Gata3 genes, each sample {.tabset}
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
for (sample.id in unique(s.obj$name)){
  cat(sprintf("### Sample %s \n", sample.id))
  p <- VlnPlot(object = subset(s.obj, name == sample.id), features = c("Ciita", "Foxp3", "Klrg1", "Gata3"), group.by = "cell.annotation", ncol = 2, assay = "SCT", slot = "data")  
  print(p)
  cat("\n \n")
}
```


```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
if (file.exists(file.path(path.to.06.output,sprintf("Project_%s_%s.rds", PROJECT, sub.cluster.idx))) == FALSE){
  saveRDS(s.obj, file.path(path.to.06.output,sprintf("Project_%s_%s.rds", PROJECT, sub.cluster.idx)))  
} else {
  s.obj <- readRDS(file.path(path.to.06.output,sprintf("Project_%s_%s.rds", PROJECT, sub.cluster.idx)))
}
```

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
#### CONVERT seurat object to cloupe file
if ("loupeR" %in% installed.packages() == FALSE){
  install.packages("hdf5r")
  install.packages("/media/hieunguyen/HD01/storage/offline_pkgs/loupeR_Linux.tar.gz", repos = NULL, type = "source")
}

if (file.exists(file.path(path.to.06.output, sprintf("PROJECT_%s_%s_cloupe_converted_from_seurat", PROJECT, 
                          sub.cluster.idx))) == FALSE){
  library(loupeR)
  loupeR::setup()
  create_loupe_from_seurat(
      s.obj,
      output_dir = file.path(path.to.06.output),
      output_name = sprintf("PROJECT_%s_%s_cloupe_converted_from_seurat", PROJECT, 
                            sub.cluster.idx),
      dedup_clusters = FALSE,
      executable_path = NULL,
      force = TRUE)  
}

```

