---
title: "Preliminary analysis, DCs clusters, extract Cd3g cells only"
author:
  - "trnguyen@ukaachen.de"
date: "Last update on `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    keep_md: no
    df_print: paged
    toc: true
    toc_float:
      toc_collapsed: false
    toc_depth: 3
    number_sections: true
    theme: lumen
params:
  PROJECT: NA
---
  
  
```{css zoom-lib-src, echo = FALSE}
script src = "https://ajax.googleapis.com/ajax/libs/jquery/3.4.1/jquery.min.js"
```

```{js zoom-jquery, echo = FALSE}
$(document).ready(function() {
  $('body').prepend('<div class=\"zoomDiv\"><img src=\"\" class=\"zoomImg\"></div>');
  // onClick function for all plots (img's)
    $('img:not(.zoomImg)').click(function() {
      $('.zoomImg').attr('src', $(this).attr('src')).css({width: '100%'});
      $('.zoomDiv').css({opacity: '1', width: 'auto', border: '1px solid white', borderRadius: '5px', position: 'fixed', top: '50%', left: '50%', marginRight: '-50%', transform: 'translate(-50%, -50%)', boxShadow: '0px 0px 50px #888888', zIndex: '50', overflow: 'auto', maxHeight: '100%'});
    });
    // onClick function for zoomImg
    $('img.zoomImg').click(function() {
      $('.zoomDiv').css({opacity: '0', width: '0%'}); 
    });
  });
```

<style type="text/css">
    div.datatables { height: auto !important;}
</style>

**Note on the .html report: Click on the image to zoom-in and one more click to go back to the original**

```{r echo=FALSE, warning=FALSE, message=FALSE, results='hide', fig.height=10, fig.width=18}
##### clean up #####
# gc()
# rm(list = ls())

path.to.project.src <- "/home/hieunguyen/CRC1382/src_2023/SBharadwaj/deep_seq_batch"

source(file.path(path.to.project.src, "00_import_libraries.R"))
source(file.path(path.to.project.src, "00_helper_functions.R"))

#####----------------------------------------------------------------------#####
# CONFIGURATIONS AND PREPRATIONS
#####----------------------------------------------------------------------#####
# outdir <- "/home/hieunguyen/CRC1382/outdir"
outdir <- "/media/hieunguyen/HD01/outdir/CRC1382/SBharadwaj_20240318"
PROJECT <- "SBharadwaj_20240318_Sample_1_4_7_8_2_5"

path.to.main.input <- file.path(outdir, PROJECT)

path.to.main.output <- file.path(path.to.main.input, "data_analysis")
dir.create(path.to.main.output, showWarnings = FALSE, recursive = TRUE)

path.to.01.output <- file.path(path.to.main.output, "01_output", "custom_output", "DCs_Cd3g")
dir.create(path.to.01.output, showWarnings = FALSE, recursive = TRUE)

path.to.07.output <- file.path(path.to.main.output, "07_output")

path.to.s.obj <- file.path(path.to.07.output, "v0.1", "DC", "s8_output/SBharadwaj_20240318_Sample_1_4_7_8_2_5_DC.output.s8.rds")

s.obj <- readRDS(path.to.s.obj)

Cd3g.cells <- GetAssayData(object = s.obj, assay = "SCT", slot = "data")["Cd3g", ]

remove.cells <- Cd3g.cells[Cd3g.cells > 0] %>% names()
keep.cells <- setdiff(colnames(s.obj), remove.cells)
```


```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.height=10, fig.width=18}
DimPlot(object = s.obj, reduction = "cca_UMAP", cells.highlight = keep.cells, label = TRUE, label.box = TRUE, repel = TRUE) +
  theme(legend.position = "none") + 
  ggtitle("UMAP: highlight cells not expressing Cd3g")
```


```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.height=10, fig.width=18}
DimPlot(object = s.obj, reduction = "cca_UMAP", cells.highlight = remove.cells, label = TRUE, label.box = TRUE, repel = TRUE) +
  theme(legend.position = "none") + 
  ggtitle("UMAP: highlight cells expressing Cd3g")
```

```{r echo=FALSE, warning=FALSE, message=FALSE, results='hide', fig.height=10, fig.width=18}
s.obj <- subset(s.obj, cells = keep.cells)

if (file.exists(
  file.path(path.to.01.output, "s8_output", sprintf("%s_DCs_Cd3g.output.s8.rds", PROJECT))
) == FALSE){
  scrna_pipeline_src <- "/home/hieunguyen/CRC1382/src_2023/src_pipeline/scRNA_GEX_pipeline_SeuratV5/processes_src"
  source(file.path(scrna_pipeline_src, "s8_integration_and_clustering_SeuratV5.R"))
  
  num.PCA <- 30
  num.PC.used.in.UMAP <- 30
  num.PC.used.in.Clustering <- 30
  num.dim.integration <- 30
  num.dim.cluster <- 30
  cluster.resolution <- 0.5
  use.sctransform <- TRUE
  vars.to.regress <- c("percent.mt")
  k.weight <- 50
  
  DefaultAssay(s.obj) <- "RNA"
  s.obj <- JoinLayers(s.obj)
  s.obj <- s8.integration.and.clustering_V5(
     s.obj = s.obj, 
     save.RDS.s8 = TRUE,
     path.to.output = path.to.01.output,
     use.sctransform = TRUE,
     num.PCA = num.PCA,
     num.PC.used.in.UMAP = num.PC.used.in.UMAP,
     num.PC.used.in.Clustering = num.PC.used.in.Clustering,
     cluster.resolution = cluster.resolution,
     vars.to.regress = vars.to.regress,
     PROJECT = sprintf("%s_DCs_Cd3g", PROJECT),
     k.weight = k.weight)
} else {
  s.obj <- readRDS(file.path(path.to.01.output, "s8_output", sprintf("%s_DCs_Cd3g.output.s8.rds", PROJECT)))
}


```


```{r echo=FALSE, warning=FALSE, message=FALSE, results='hide', fig.height=10, fig.width=18}
Idents(s.obj) <- "cca.cluster.0.5"
if (file.exists(file.path(path.to.01.output, "DE_cluster_marker_genes.rds")) == FALSE){
  DefaultAssay(s.obj) <- "SCT"
  s.obj <- PrepSCTFindMarkers(s.obj)
  cluster.markers <- FindAllMarkers(object = s.obj, assay = "SCT", test.use = "wilcox", slot = "data", min.pct = 0.5)
  cluster.markers <- subset(cluster.markers, cluster.markers$p_val_adj < 0.05 & cluster.markers$avg_log2FC > 0)
  saveRDS(cluster.markers, file.path(path.to.01.output, "DE_cluster_marker_genes.rds"))
} else {
  cluster.markers <- readRDS(file.path(path.to.01.output, "DE_cluster_marker_genes.rds"))
}

```

# Dimension reduction with UMAP

## UMAP: all clusters 
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
DimPlot(object = s.obj, reduction = "cca_UMAP", label = TRUE, label.box = TRUE, label.size = 10, pt.size = 1, repel = TRUE, group.by = "cca.cluster.0.5") + 
  ggtitle(sprintf("UMAP: All clusters")) +
  theme(axis.text = element_text(size=20, face = "bold"),
        axis.title = element_text(size=20, face = "bold"), 
        title = element_text(size = 20, face = "bold"))
```

## UMAP: all samples
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
DimPlot(object = s.obj, reduction = "cca_UMAP", label = TRUE, label.box = TRUE, label.size = 10, pt.size = 1, repel = TRUE, group.by = "name") + 
  ggtitle(sprintf("UMAP Sample: All samples after integrating")) +
  theme(axis.text = element_text(size=20, face = "bold"),
        axis.title = element_text(size=20, face = "bold"), 
        title = element_text(size = 20, face = "bold"))
```



# Cluster marker genes

Identify differentially expressed genes in each cluster. 

## Feature plot {.tabset}
```{r echo=FALSE, fig.height=12, fig.width=20, message=FALSE, warning=FALSE, results='asis'}
for (cluster.id in sort(unique(cluster.markers$cluster))){
  DefaultAssay(s.obj) <- "SCT"
  cat(sprintf("### Cluster %s \n", cluster.id))
  tmp.cluster.markers <- subset(cluster.markers, cluster.markers$cluster == cluster.id) %>% arrange(desc(avg_log2FC))
  p <- FeaturePlot(object = s.obj, reduction = "cca_UMAP", features = head(tmp.cluster.markers, 9)$gene, ncol = 3, label = TRUE, pt.size = 0.5, label.size = 5, label.color = "red", slot = "data")  
  print(p)
  cat("\n \n")
}
```

## Dot plot{.tabset}
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
for (cluster.id in sort(unique(cluster.markers$cluster))){
  DefaultAssay(s.obj) <- "SCT"
  cat(sprintf("### Cluster %s \n", cluster.id))
  tmp.cluster.markers <- subset(cluster.markers, cluster.markers$cluster == cluster.id) %>% arrange(desc(avg_log2FC))
  p <- DotPlot(object = s.obj, features = head(tmp.cluster.markers, 9)$gene)  
  print(p)
  cat("\n \n")
}
```

## Violin plot{.tabset}
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
for (cluster.id in sort(unique(cluster.markers$cluster))){
  DefaultAssay(s.obj) <- "SCT"
  cat(sprintf("### Cluster %s \n", cluster.id))
  tmp.cluster.markers <- subset(cluster.markers, cluster.markers$cluster == cluster.id) %>% arrange(desc(avg_log2FC))
  p <- VlnPlot(object = s.obj, features = head(tmp.cluster.markers, 9)$gene, slot = "data")  
  print(p)
  cat("\n \n")
}
```

## Full tables of DE genes {.tabset}
```{r echo=FALSE, results='asis', include=FALSE}
cluster.markers %>% create_dt()
```

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
all.cluster.ids <- sort(unique(s.obj$cca.cluster.0.5))
for (cluster.id in all.cluster.ids){
  tmp.table <- subset(cluster.markers, cluster.markers$cluster == cluster.id) %>% arrange(desc(avg_log2FC))
  cat(paste("\n\n### Cluster: ", cluster.id, "##\n"))
  print( htmltools::tagList(datatable(tmp.table, extensions = 'Buttons',
                                      filter = "top",
                                      options = list(dom = 'Blfrtip',
                                                     buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
                                                     lengthMenu = list(c(10,25,50,-1),
                                                                       c(10,25,50,"All")),
                                                     columnDefs = list(list(
                                                       targets = "_all",
                                                       render = JS(
                                                         "function(data, type, row, meta) {",
                                                         "return type === 'display' && data != null && data.length > 100 ?",
                                                         "'<span title=\"' + data + '\">' + data.substr(0, 100) + '...</span>' : data;",
                                                         "}")
                                                     ))
                                      ))))
  cat("\n \n")  
}
```

# Number of cells in each cluster in each sample
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
cell.count.in.clusters <- table(s.obj$name, s.obj$cca.cluster.0.5) %>% as.data.frame() %>%
  pivot_wider(names_from = "Var2", values_from = "Freq") 
cell.count.in.clusters %>% create_dt()
```

# Cell type annotation by pathway analysis with CellMarkers 2.0
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
if (packageVersion("clusterProfiler") != "4.10.1"){
  remove.packages("clusterProfiler")
  remove.packages("DOSE")
  remove.packages("GOSemSim")
  path.to.install.dir <- "/media/hieunguyen/HD01/storage/offline_pkgs/clusterProfiler"
  install.packages(file.path(path.to.install.dir, "HDO.db_0.99.1.tar.gz"), type = "source", repos = NULL)
  install.packages(file.path(path.to.install.dir, "yulab.utils_0.1.4.tar.gz"), type = "source", repos = NULL)
  install.packages(file.path(path.to.install.dir, "GOSemSim_2.28.1.tar.gz"), type = "source", repos = NULL)
  install.packages(file.path(path.to.install.dir, "DOSE_3.28.2.tar.gz"), type = "source", repos = NULL) 
  install.packages(file.path(path.to.install.dir, "gson_0.1.0.tar.gz"), type = "source", repos = NULL)
  install.packages(file.path(path.to.install.dir, "clusterProfiler_4.10.1.tar.gz"), type = "source", repos = NULL) 
}

library(clusterProfiler)
```

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
if ("org.Mm.eg.db" %in% installed.packages() == FALSE){
  BiocManager::install("org.Mm.eg.db")
}
library("org.Mm.eg.db")

annotate_with_pathway <- function(cluster.id, topN = 20){
  if (topN == "all"){
      tmpdf <- subset(cluster.markers, cluster.markers$cluster == cluster.id) %>% arrange(desc(avg_log2FC))
  } else {
      tmpdf <- subset(cluster.markers, cluster.markers$cluster == cluster.id) %>% arrange(desc(avg_log2FC)) %>% head(topN)    
  }

  path.to.cellmarker2.0 <- "/media/hieunguyen/HD01/storage/CellMarker2.0"
  cell_marker_data <- readxl::read_excel(file.path(path.to.cellmarker2.0, "Cell_marker_Mouse.xlsx"))
  
  ## instead of `cellName`, users can use other features (e.g. `cancerType`)
  cells <- cell_marker_data %>%
      dplyr::select(cell_name, GeneID) %>%
      dplyr::mutate(GeneID = strsplit(GeneID, ', ')) %>%
      tidyr::unnest()
  
  convertdf <- bitr(tmpdf$gene, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Mm.eg.db")
      
  x <- enricher(convertdf$ENTREZID, TERM2GENE = cells) %>% 
    as.data.frame() %>%
    subset(p.adjust <= 0.05)
  x <- x %>% subset(select = -c(geneID)) %>% arrange(desc(Count))
  return(x)  
}

pathway.annotation <- hash()
```

## Result tables {.tabset}
```{r echo=FALSE, results='asis', include=FALSE}
annotate_with_pathway(1) %>% create_dt()
```

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
annotationdf <- data.frame()
top5.annotationdf <- data.frame()

available.diff.clusters <- unique(cluster.markers$cluster)

for (cluster.id in all.cluster.ids){
  if (cluster.id %in% available.diff.clusters == FALSE){
    tmp.table <- data.frame(status = c("Cell type not detected!"))
  } else {
      tmp.table <- annotate_with_pathway(cluster.id)
  }

  if (nrow(tmp.table) == 0 | "status" %in% colnames(tmp.table)){
    tmp.table <- data.frame(status = c("Cell type not detected!"))
  } else {
      tmpdf <- data.frame(cluster.id = c(cluster.id), annotation = c(head(tmp.table, 1)$ID))
      annotationdf <- rbind(annotationdf, tmpdf)
      
      tmpdf <- data.frame(cluster.id = c(cluster.id), annotation = c(paste(head(tmp.table, 5)$ID, collapse = ", ")))
      top5.annotationdf <- rbind(top5.annotationdf, tmpdf)
    
  }
  
  cat(paste("\n\n### Cluster: ", cluster.id, "##\n"))
  print( htmltools::tagList(datatable(tmp.table, extensions = 'Buttons',
                                      filter = "top",
                                      options = list(dom = 'Blfrtip',
                                                     buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
                                                     lengthMenu = list(c(10,25,50,-1),
                                                                       c(10,25,50,"All")),
                                                     columnDefs = list(list(
                                                       targets = "_all",
                                                       render = JS(
                                                         "function(data, type, row, meta) {",
                                                         "return type === 'display' && data != null && data.length > 100 ?",
                                                         "'<span title=\"' + data + '\">' + data.substr(0, 100) + '...</span>' : data;",
                                                         "}")
                                                     ))
                                      ))))
  cat("\n \n")  
}
```

## Top-1 annotation for each cluster
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
annotationdf %>% create_dt()
```

## Top-5 annotation for each clusters
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
top5.annotationdf %>% create_dt()
```

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
meta.data <- s.obj@meta.data %>% rownames_to_column("barcode") %>%
  rowwise() %>%
  mutate(prediction2 = ifelse(nrow(subset(annotationdf, annotationdf$cluster.id == cca.cluster.0.5)) != 0, 
                              subset(annotationdf, annotationdf$cluster.id == cca.cluster.0.5)$annotation,
                              "None")) %>%
  column_to_rownames("barcode")
meta.data <- meta.data[row.names(s.obj@meta.data),]
s.obj <- AddMetaData(s.obj, metadata = meta.data$prediction2, col.name = "prediction2")

DimPlot(object = s.obj, reduction = "cca_UMAP", label.box = TRUE, label = TRUE, repel = TRUE, group.by = "prediction2")
```

## Number of cells in each annotated cluster (top-1 annotation)
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
table(s.obj$prediction2, s.obj$name) %>% as.data.frame() %>% pivot_wider(names_from = "Var1", values_from = "Freq") %>% create_dt()
```

# UMAP: expression of some genes

## All samples in integrated UMAP {.tabset}
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
Idents(s.obj) <- "cca.cluster.0.5"
FeaturePlot(s.obj, features = c("Ciita", "Foxp3", "Klrg1", "Gata3"), label = TRUE, reduction = "cca_UMAP")
```

## Each sample in integrated UMAP {,tabset}
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
Idents(s.obj) <- "cca.cluster.0.5"
for (sample.id in unique(s.obj$name)){
  cat(sprintf("### Sample %s \n", sample.id))
  p <- FeaturePlot(subset(s.obj, name == sample.id), features = c("Ciita", "Foxp3", "Klrg1", "Gata3"), label = TRUE, reduction = "cca_UMAP")  
  print(p)
  cat("\n \n")
}

```

## Violin plot: Ciita, Foxp3, Klrg1, Gata3 genes, all samples

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
VlnPlot(object = s.obj, features = c("Ciita", "Foxp3", "Klrg1", "Gata3"), group.by = "cca.cluster.0.5", ncol = 2, assay = "SCT", slot = "data")
```

## Violin plot: Ciita, Foxp3, Klrg1, Gata3 genes, each sample {.tabset}
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
for (sample.id in unique(s.obj$name)){
  cat(sprintf("### Sample %s \n", sample.id))
  p <- VlnPlot(object = subset(s.obj, name == sample.id), features = c("Ciita", "Foxp3", "Klrg1", "Gata3"), group.by = "cca.cluster.0.5", ncol = 2, assay = "SCT", slot = "data")  
  print(p)
  cat("\n \n")
}
```


```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
if (file.exists(file.path(path.to.01.output,sprintf("Project_%s.rds", PROJECT))) == FALSE){
  saveRDS(s.obj, file.path(path.to.01.output,sprintf("Project_%s.rds", PROJECT)))  
} else {
  s.obj <- readRDS(file.path(path.to.01.output,sprintf("Project_%s.rds", PROJECT)))
}
```

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
#### CONVERT seurat object to cloupe file
# if ("loupeR" %in% installed.packages() == FALSE){
#   install.packages("hdf5r")
#   install.packages("/media/hieunguyen/HD01/storage/offline_pkgs/loupeR_Linux.tar.gz", repos = NULL, type = "source")
# }
# 
# library(loupeR)
# 
# loupeR::setup()
# create_loupe_from_seurat(
#     s.obj,
#     output_dir = file.path(path.to.01.output),
#     output_name = sprintf("PROJECT_%s_cloupe_converted_from_seurat", PROJECT),
#     dedup_clusters = FALSE,
#     executable_path = NULL,
#     force = TRUE
#     )
```


